|

### 2019-12-25 article/letter.Rmd
add letter, edit appendix

\noindent 2 December 2019

@@

We have worked specially hard to clarify the connection between information provision and mobilisation, a point emphasised by both the editor and the reviewer. Our efforts in that regard include, first, a new section in the SI Appendix with a detailed description of _Tá de Pé_'s functionalities and the associated Facebook campaigns. We explain the precautions we have taken to ensure that we correctly identify the effect of the TDP app on school outcomes. Second, we provide a thorough discussion of our additional data sources. More specifically, we show how the Brazilian Ministry of Education collects and publicises their data and indicate how our experiment uses that information to evaluate the robustness of our findings. We believe the new sections help give the results a clearer interpretation and strenghten our main arguments.

Related, we have included a discussion regarding app usage as suggested by the reviewer. We have included information provided by Google Analytics not only on app downloads but also on app usage, which addresses another important concern raised by the reviewer. The data show that the app usage was significant, yet it was not enough to trigger any responses from local governments.

Lastly, we have rewritten some parts of the main text to highlight that our experiment hinges on participation only, and that the effect of mobilisation and pressure on school construction outcomes require further analyses. We believe our paper now reflects our results more accurately.

@@

\noindent Sincerely,

@@

**1)** The editor writes:

\noindent 
**Response:**

\begin{quotation}
Lorem ipsum.
\end{quotation}

@@

**1)** Reviewer 1 writes:

\noindent
**Response:**

\newpage
\setlength{\parindent}{0cm}
\setlength{\parskip}{5pt}

>>>>

\noindent 25 December 2019

@@

We have worked specially hard to clarify the connection between information
provision and mobilisation, a point emphasised by both the editor and the
reviewer. Our efforts in that regard include, first, two sections in the online
appendix with a detailed description of _Tá de Pé_'s functionalities and the
associated Facebook campaigns. We explain the precautions we have taken to
ensure that we correctly identify the effect of the TDP app on school outcomes.
Second, we provide a thorough discussion of our additional data sources in
sections 3 and 4 of the appendix. More specifically, we show how the Brazilian
Ministry of Education collects and publicises their data and how we use that
information to evaluate the robustness of our findings. We believe the new
sections help give the results a clearer interpretation and strenghten our main
arguments.

Related, we have included information on app use and app downloads, which
addresses another important concern raised by the reviewer. The data have been
provided by Google Analytics and show that app usage was significant during the
two interventions, yet the usage was not sufficient to trigger any responses
from local governments.

Lastly, we have rewritten some parts of the main text to highlight two points:
1) that our experiment hinges on participation only; 2) that the effect of
mobilisation and pressure on school construction outcomes require further
analyses. We believe our paper now reflects our results more accurately.

@@

\vspace{.5cm} \noindent Sincerely,

@@

\textbf{1)} The editor writes: "_The reviewer confirmed my own sense that the
manuscript and underlying field experiment are interesting, although you will
see that the reviewer suggests many ways the manuscript can be improved.  I am
in full agreement with the reviewer that you need to reflect more deeply about
the theory and evidence for why information provision via the app could be
connected to mobilization, -- otherwise the null result is not interesting. As
the reviewer notes, the null result could be due to a disconnect between
information provision and action, or due to who selects into downloading the
app._"

\vspace{.5cm} \noindent \textbf{Response:}

> We agree that the link between information provision and mobilisation was not
clearly defined in the original manuscript. We believe the main issue lies in
the lack of evidence for mobilisation, that is, we did not provide sufficient
information about how users interacted with the application and whether they
have used the app consistently. In the revised version of the manuscript, we
provide data from Google Analytics which indicate that users did engage with
the app. According to the service, the app had 6,092 active users in
intervention 1 and gained 4,078 new users during intervention 2. On average,
each user launched 60 sessions per app install, an indication of their actual
engagement with the application. Moreover, the data also show that the app had
more than 53,000 screen visualisations, with an average of 2.42 visualisations
per user. We have added this information to the main text and to the online
appendix. The additions are marked in \textbf{boldface}. We elaborate further
on this issue in the response to Reviewer 1 below.

**1)** Reviewer 1 writes: "[...] _My biggest concern is there is no discussion
of the actual bottom-up accountability that may (or may not) have occurred._"

\vspace{.5cm} \noindent \textbf{Response:}

> We have made several changes to the main paper and the online appendix to
address this important concern. First, we have added a new section to the
appendix (B) in order to clarify how we designed and conducted the experiment.
We copy the text below:

@@

> **Appendix C -- Treatment Definition and Mechanism**

> _In this section, we explain in detail how the _Tá de Pé_ app works. The
treatment involves the following procedures:_

> _1. Users download the app._

> _2. Users access the app and add new information about school constructions
	 near to their location._

> _3. The app sends a message to the mayor's office requesting details about
the project. If necessary, the message is also forwarded to other government
oversight institutions such as the city council and the Brazilian Federal
Comptroller (CGU). Below, we discuss the cases in which we forwarded the
denounce to the authorities._

> _4. If the pressure over the mayors has an effect, we expect to witness
	 improvements on governmental data regarding the school construction
	 progress._

> _The treatment is a compounded treatment comprised of steps (2) to (4). To
make sure that the only source of effect comes from step (2), we took the
following precautions:_

> _In step (1), we have implemented Facebook campaigns in municipalities that
have school construction agreements with the Federal government. We did not
discriminate any municipality in this step, and citizens living in
municipalities in the treatment and in the control could equally have received
the banners about the _Tá de Pé_ app. The Facebook campaign was not
randomised._

> _For step (3), we put a team of two data scientists to ensure that upon
receiving a denounce, the mayor's office, the city council, and the Brazilian
Federal Comptroller (CGU) would be notified. The decision making funnel was:_

> _3.1. We notified the _Engineers without Borders_ (EWB) about the denounce
and sent them the user-provided pictures and auxiliary information. They issued
a report on how much the construction was likely to be delayed._

> _3.2. After we had the EWB report, we notified the mayor's office in
compliance with Brazil's Freedom of Information Act (FOI). We communicated that
a citizen had anonymously made a denounce about a school (fully identifying the
school construction that we were referring to), and that the Engineers without
Borders found the construction delayed. We submitted a request for the mayor's
office to provide an explanation on why the construction was delayed, and asked
for updated statistics on when the construction will be finished. According to
the FOI law, the mayor had twenty days to respond._

> _3.3. If the response was deemed insufficient, we forwarded the denounce to
the city council and to the Brazilian Federal Comptroller (CGU). As the
denounce goes to these institutions, they can harm mayors by restricting access
to Federal funds._

> _3.4. At each stage, a Twitter bot constructed by Transparência Brasil
(<https://twitter.com/tadepeapp>) reported the reception of a denounce and
tracked the responses of the mayor, the city council, and the CGU._

> _In step (4), we relied on administrative data by the Brazilian Ministry of
Education. The Ministry of Education provides monthly data on school
constructions build with Federal Government resources and shows how many of
them are delayed across the country. We use the finishing dates reported by the
municipality at the time they sign the agreement with the Federal Government as
our measure of construction delays. This data helped us to keep the app
updated. We selected outcomes six months before and after the treatment. The
data were constantly updated by the Ministry of Education, and we found no red
flags in the overall quality of their information._

> _Therefore, steps (1), (3), and (4) were standardised and do not vary across
the country. The only source of variation in our intervention should come from
step (2)._

> _In step (2), we applied the treatment by removing schools within
municipalities out of the app. In the first intervention, which happened at the
municipal level, we excluded all schools within a given randomly selected
municipality. In the second intervention, we randomly excluded selected schools
out of the app. Removing a school means that the school was in the dataset
provided by the Brazilian Ministry of Education, but it did not show up in the
app. Therefore, our treatment effect represents the school being present
(treatment) or absent (control) in the app._

> _We argue that the app promotes two crucial features of bottom-up
accountability. First, it helps people to access information that would
otherwise be available only in an obscure spreadsheet on the Ministry of
Education website. Second, whenever people find this information troublesome,
we are providing an easy way to them to act upon the information. As the _Tá de
Pé_ design was intuitive and straightforward, we believe that most of
Brazilians would find it easy to use the app._

> _Moreover, the success of treatment does not require that everyone in the
municipality uses it. The more usage means more pictures and denounces, and
this represents more pressure over the mayors office, what could revert into a
higher effect. However, just one denounce would be enough for the project to be
successful, as the denounce can be directed to the Brazilian General
Comptroller (CGU), and this could potentially deny future funding to the
municipality._

\noindent \textbf{2)} Reviewer 1 writes: "_The authors have devised a nice
experiment and I believe their design to have been appropriately administrated.
But the design hinges on several critical mechanisms that go largely
undiscussed: not only must information be available to citizens such that they
might participate in local oversight, but said citizens must 1) access the
information and 2) act on said information. This is mentioned in the
Discussion---that perhaps the app was insufficient to spur collective
action---but that also seems like an empirical question the authors could
feasibly address_. [...] _Once people downloaded the App, did they use it?_"

> To answer these questions, we have added Google Analytics data on _Tá de
Pé_'s download and usage. This provide reliable and exogenous data
about our experimental manipulation and the effectiveness of the treatment. The
data confirm that app usage was significant. We add the following paragraph to
the main text to show user engagement (page 8):

> _Data from Google Analytics suggest that users did engage with the TDP app.
On average, each user launched 59.99 app sessions, which indicates their
interest in the application. In total, the app had 53,928 screen
visualisations, with an average of 2.42 screen visualisations per session._

> We see that users not only downloaded the app but visualised
different screens at each session. This means that users did check all pages of
the app, including those with information about local school constructions.
Moreover, figures 2 and 3 in the main text (pages 7 and 8) also show that the
downloads were widespread across the country. Considering Brazil's large
territory, we believe this is robust evidence of an effective treatment
manipulation.

> As we note in Appendix C, one denounce would be sufficient for the city
council and the Federal Comptroller to start an investigation against the
offending mayor. However, we do not find any substantial treatment effect in
our two interventions. We have added three paragraphs to the Discussion section
on possible explanations for the null effects (pages 12 and 13):

> _What factors, then, are driving these results? It seems unlikely that the
null results derive from flaws in the research design. First, our study is well
powered. Although the treatment is indirect---the person has to download the
app, find a school construction, and then report it---,we included a
substantial number of schools in the treatment groups. Second, balance and
manipulation tests indicate that the treatment allocation was successful, so we
can rule out problems in the randomisation procedures. Third, data from
Google Analytics confirm that citizens did use the app and provided information
to our dataset. This indicates that the treatment manipulation was effective.
In this sense, it is unlikely that our results derive from low user response.
Fourth, after doing a series of robustness tests, we still find no firm
evidence of treatment effect. Finally, note that the sign of the coefficients
are frequently contrary to our theoretical expectations. This rules out a
possible concern about statistical power with our small control group
approach._

> _We discuss some possible reasons why community monitoring did not work in
our case. One plausible explanation is that individuals were unable to
differentiate the effect of political corruption from those of spending cuts.
Due to the severe economic crisis in 2014--2016, the Brazilian federal
government introduced discretionary spending limits that affected public
investment [@rossi2016impactos]. Politicians may argue that delays in school
constructions are not derived from their misuse of government funds, but from
the austerity measures. If this is the case, citizens will not blame local
politicians for the underprovision of public goods. Consequently,
representatives can dismiss individual requests as the issue is unlikely to
escalate._

> _The electoral cycle might also have decreased the potential effect of the
treatment. As the experiment was fielded right after Brazil's municipal
elections, incumbents might have disregarded the requests because they did not
see the demands as politically costly in the short run. Having just taken
office, mayors might have focused their attention to the formation of
government coalitions or to budget concerns. Future research may evaluate how
electoral dynamics interact with citizen oversight, potentially by replicating
informational experiments in different stages of the political cycle._

\noindent \textbf{3)} Reviewer 1 writes: "_Did the creators of the App
actually gather information on school construction?_"

> We have added a new section to the online appendix describing our data
collection process (Appendix D). The section contains our data sources and a
brief description of each outcome variables we employed in the paper. The
variables we use were compiled by the Brazilian Ministry of Education. The new
section is available below.

> **Appendix D -- Outcome Data Collection**

> _We collect the main outcomes using the data reported by the municipality to
the Brazilian Ministry of Education. Note that this is the same data as the
pre-treatment data that we use to feed in the app with delayed schools._

> _The outcomes used in the paper are:_

> _1. **Percentage of the project completed before the impact evaluation
	 started**: This is a column in the Ministry of Education dataset. The data
	 in this column were sampled in the beginning of the intervention._

> _2. **Percentage of the project reported as completed by the end of the
	 intervention period**: Another column in the Ministry of Education dataset.
	 The data were sampled six months after the intervention._

> _3. **Difference between the percentage reported as completed before and
after the intervention**: The difference between (1) and (2) above. Note that
according to @gerber2012field, this way to look into outcomes is positive, as
having a pre-treatment covariate that highly predicts the outcome reduces the
variance in the estimator considerably._

> _4. **Dummy indicator for finished constructions**: The Ministry of Education
	 dataset has a column with the construction status. We created a dummy to be
	 one when the construction was reported as finished and zero otherwise. The
	 data here were sampled six months after the intervention._

> _5. **Dummy indicator for cancelled constructions**: Similarly, we created a
	dummy variable from the Ministry of Education dataset indicating when the
	construction was reported as cancelled. The data here were sampled six months
	after the intervention._

> _6. **Number of schools where construction companies updated the conclusion
	 dates**: Lastly, the Ministry of Education dataset has a column with the
	 construction estimated completion date. We took the completion dated
	 reported before and six months after the intervention. If the dates
	 differed, we coded as one, and zero otherwise._

> _All these variables come from our manipulation of the Ministry of Education
data. Collecting the outcomes in this way is superior than the alternative of
gathering the data ourselves. First, we would not be able to manipulate the
results and cherry-pick favourable results. Even if they have the best
intentions, researchers may bias the results just by the choices they make in
the data collection process. For instance, if we send an RA to collect the
data, as she knows the app, she could unconsciously over-report positive
results. This would bias our findings and undermine the credibility of our
experiment. Second, using external data makes the results easily reproducible
by other researchers. As the Ministry of Education has standardised data
collection procedures, they greatly simplify and facilitate the reproduction of
our findings._

\noindent \textbf{4)} Reviewer 1 writes: "_Some of this basic information is
fascinating, and two critical components to how we ought to interpret the
results. For example, page 8 "The literature on bottom up accountability argues
that more information about the shortcomings of public service provision will
mobilise citizens who, in turn, would pressure state agents to produce better
social policies. Our results do not lend support to that hypothesis." Your
results suggest that providing information alone is not sufficient, but
different analyses are required to evaluate the mobilization and pressure part
of that equation._"

> We fully agree with the reviewer and have changed the main text accordingly.
We have rewritten several passages, from the abstract to the discussion, in
order to highlight that our mechanism is indeed based on information provision
and not exactly mobilisation. For instance, the first paragraph in the
Discussion session has been rewritten as:

> _In this paper, we discuss **whether delivering information to citizens via** a
mobile phone application fosters ~~grassroots mobilisation~~ **community
oversight** and political accountability in Brazil. Our two interventions show
that the results are at best mixed. Although we find ~~an~~ **some treatment**
effect on school cancelling rates in the first intervention, the app has no
consistent impact on our outcomes of interest. These findings add to the
studies that cast doubts on the relationship between bottom-up accountability
and local policy performance [e.g., @banerjee2010pitfalls; @lieberman2014does;
@raffler2018weakness]._

Similarly, we have changed the Introduction to:

> _We use the TDP app to conduct two experimental interventions and test ~~its
impact~~ **the impact of citizen oversight** on five outcomes related to school
completion rates and complaints to public authorities. Overall, ~~our results
show that~~ providing information to citizens has no consistent impact on
policy outcomes._ [...] _Our results are in line with the
latter group and suggest that local oversight is ineffective in altering
government behaviour in Brazil._

\noindent \textbf{5)} Reviewer 1 writes: "_A similar short coming is the lack
of discussion of who downloaded the App, and if/how said characteristics
correlate with municipal level predictors of school construction outcomes. It
might be that the randomization "worked" at the municipal level (balance and
manipulation tests suggests this is the case), but if the App was exclusively
downloaded by teenagers (those most likely to be on Facebook, but least likely
to care about public works information, and least likely to do something about
it), then that could be the root of the problem. Indeed, the authors describe
in the SI Appendix the ability to evaluate subsets of user data, I'm curious as
to why they did not._"

> The reviewer correctly points out that the usage of the app could have
suffered from heterogeneous effects on the types of users achieved by the
Facebook campaign. To clarify this point, first, we would like to note that the
Facebook campaign was not randomised. This means that every municipality that
had a delayed construction site in the Ministry of Education data was targeted,
regardless of being in the treatment or the control group. In fact, the app
could be downloaded in any Brazilian municipality, regardless of having a
construction delayed on it. The App will automatically list the school sites
closer to the user, even if the closest school is kilometres away from the
user.

> Second, the Brazilian countryside is still plagued by political violence
against citizens acting against mayors. Because of this possibility, we decided
to make the app completely anonymous. Although this carries a cost in terms of
the analysis, it ensures that the citizens were not exposed in any way.
Moreover, the randomization is known for washing out any the interference of
any unobserved factor that could threat the identification of the causal
effect. The success on observables reassures us that the randomisation was well
performed, and probably eliminated the concern of heterogeneity from
non-observables. However, we agree with the reviewer that the target public
could be either outside Facebook, or being of a given demographic
characteristic that we did not captured in our experiment.

> We have also included further information on the Facebook campaign in the
> Appendix (section E):

> **Appendix E: Facebook Tá de Pé Campaign**

> _The app by itself would probably not have had an effect on the
responsiveness of mayors. To boost usage and publicize the app, Transparência
Brasil ran 64 Facebook campaigns throughout the country. The Facebook campaign
concentrates the bulk of the investments in the beginning of the intervention.
The app download data is in the second panel of Figures 2 and 3. For the
Facebook interaction, we have the following statistics:_

> _1. Impressions_
	- _Mean: 36052.69_
	- _Total for all campaigns: 5,155,535_

> _2. App Download link clicks:_
	- _Mean:  24398.65_
	- _Total for all campaigns: 1,951,892

> _As we see from the actual download activity, the people who indeed
downloaded the app were much fewer than these numbers would suggest._

> _We have chosen Facebook because the platform offers an easy and cheap method
to spread information. Moreover, according to the Statista website, Facebook is
used by 70 million Brazilians (60% of the entire adult population), and the
demographics largely match the demographics publicised by the Brazilian
Institute of Geography and Statistics. For instance, see the following
websites:_

> _1. [Statista Total Facebook
> Users](https://www.statista.com/statistics/244936/number-of-facebook-users-in-brazil/)_

> _2. [Statista Total Facebook Users by
	 Age](https://www.statista.com/statistics/866282/facebook-user-share-brazil-age/)_

\noindent \textbf{6)} Reviewer 1 writes: "_In light of the aforementioned
nuance that is lacking but ought to be improved, I encourage the authors to
take care in how they describe their experiment, the treatment and what we can
infer. For example, the first sentence of the abstract asks: “Does grassroots
participation improve public service delivery?” The analysis hinges not on
participation (something on which we have no information), but the possibility
of oversight._ [...] _This is but one example, I would encourage the authors to
carefully revise an already polished manuscript with these sorts of clarifying
issues in mind._"

> We agree with the reviewer and have followed his/her suggestions for revising
the phrasing of some of our passages. We believe the text has been
significantly improved as a result. The changes can be read in several parts of
the main text, but we would like to share some examples here. For instance, the
abstract now reads:

> _Does ~~grassroots participation~~ **local oversight** improve public service
> delivery? We study the effect of a mobile phone application that allows
> citizens to monitor school construction projects in Brazilian municipalities.
> ~~The app provides a platform where users can submit photos of construction
> sites, consult independent engineers, and contact the mayor's office about
> project delays.~~ **The app prompts users to submit data about construction
> sites, sends such crowdsourced information to independent engineers, and
> contacts the mayors' offices about project delays.** Our results show that
> the app has a null impact on the school construction indicators.
> Additionally, we find that politicians are unresponsive to individual
> requests. The results question the impact of ~~local~~ **bottom-up**
> monitoring on public service performance and suggest that interventions
> targeted ~~at other groups~~ **to other groups, or focusing on different
> issues,** may produce better policy outcomes._

Our explanation of the results on page 9 has been rewritten as:

> _We find that the app only has a small effect on cancellation rates. The TDP
application increases the likelihood of cancelling the construction in 2.07
percent. While this result goes in the opposite direction of our theoretical
expectations, the finding is inconsistent and does not replicate in the second
experiment. All other coefficients are not statistically significant at
conventional levels. On the one hand, the results indicate that our placebo
outcome, the percentage of the invested executed before the intervention,
indeed behaves as predicted. On the other hand, we expected the five remaining
outcomes to improve after the introduction of the app. The literature on
bottom-up accountability argues that delivering more information about the
shortcomings of public services provision will ~~mobilise citizens who, in
turn, would pressure state agents to produce better social policies~~ **put
citizens in a position where they can monitor state agents and improve provider
behaviour** [@raffler2018weakness]. Our results do not lend support to that
hypothesis._

\noindent \textbf{7)} Reviewer 1 writes: "_Can you address possible
cross-municipality contamination?_"

> The reviewer finally questions us about the possibility of cross-municipality
> contamination. This would mean that the existence of the app in a
> municipality could make mayors in other municipalities to act and increase
> their efforts toward getting the schools finished. She/He is correct that, if
> this is the case, the null result should not be surprising. However, looking
> into changes in outcomes in the control and the treatment we see no effect of
> proximity. Moreover, municipalities in Brazil are much insulated, and mayors
> usually belong to different parties. There are about 35 active parties in
> Brazil, which makes cross-municipality contact and information sharing costly
> and difficult. Therefore, we believe that this possibility is remote.

We would like to thank Reviewer 1 for his/her helpful comments and suggestions.

\setlength{\parindent}{0cm} \setlength{\parskip}{5pt}
